Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-10-22 TACO-paper-v2/TACO-Rebuttal-final.txt

We thank our reviewers (R1-R4). We appreciate the feedback and the
critical comments on our paper obtained in the revision process, which
really led to a significantly improved version. Below we detail how we
answered the minor points raised by R1 and R4.

R1:
1) "I'm not hung up over tables II and III, but putting them in small graphs
would, I think, still be good for the final version, space permitting."

  We agree that it would probably be easier to read plots than
  tables. We will exceed the 25-page limit, if we do so. In addition,
  we feel that the best way of representing a table such as Tables II
  and III is by means of 6 plots (3 plots per architecture, each plot
  corresponding to a specific partial differential equation). Due to
  lack of space, we prefer to stick to the tables rather than adding a
  total of 12 tiny plots, which would really hinder readability.

@@

2) "Also, your comment about different compilation flags (page 13,
line 18) is still not clear, in my opinion.  It reads like other
optimization levels were worse only for the Xeon Phi. If you could
just clarify that you picked the best optimization levels for each
platform, that make your methodology more clear."

  Thanks for the suggestion. The sentence was not that clear. So we
  rephrased it as follows, which hopefully avoids any ambiguity:

    "... The icc 13.1 compiler was used. We selected the best
     optimization levels for each platform; we used -xAVX for
     auto-vectorization and -O2 on the Sandy Bridge, and -O3 on the
     Xeon Phi."

R4:

1) "Ideally, we would like COFFEE to be platform independent. However,
in this work, COFFEE is tightly coupled with the target platform. It
generates vector intrinsic and works closely with the PyOP2
runtime. It would be better if the authors could show that the
approach would also work well on a different platform, such as ARM,
SPARC, PowerPC or MIPS. Nevertheless, this is merely a suggestion, not
a requirement, since we understand how much work is involved in tuning
for a different platform."

  We emphasize that the only way in which COFFEE is tied with the
  architecture is by means of vector intrinsics (in this case, related
  to the x86 instruction set), while all of the other transformations
  occur at the source code level. Using some abstractions on top of
  vector intrinsics would probably fix this; this is something we
  planned for the near future, since Firedrake is going to be ported
  to BLUE GENE architectures. COFFEE is not tied with the PyOP2
  runtime either: it is true that a suitable interface is used in
  PyOP2 to interact with COFFEE; however, COFFEE easily can be used in
  scenarios other than Firedrake/PyOP2.

@@

2) "The new title of the paper is now changed to focus on cross-loop
optimization.  However, for modern microprocessors, cross-loop
optimization usually includes locality improvement transformations to
increase the performance of cache hierarchies. COFFEE may not need to
do that since it focuses on lower order methods where the working set
is already fit in the L1 cache. However, the paper should point out
why locality improvement is not in their scope."

  We note that locality improvement is already in the scope of the
  paper: we optimize for data locality at the level of registers,
  which is the level of the memory hierarchy immediately below the L1
  cache. In particular, the outer-product vectorization and expression
  splitting aim to optimize for register data locality. However, we
  agree with you that it is useful to remark this aspect. Therefore,
  following your suggestion, we have rewritten a few key
  sentences. Firstly, in the abstract, we changed the following
  sentence:

    "The compiler manipulates abstract syntax trees generated from a
     domain-specific language by introducing domain-aware
     optimizations and, eventually, produces C code including vector
     SIMD intrinsics. >>

@@

  into:

    "The compiler manipulates abstract syntax trees generated from a
     domain-specific language by introducing domain-aware
     optimizations for instruction-level parallelism and register
     locality. Eventually, it produces C code including vector SIMD
     intrinsics."

  This would alert the reader from the very beginning that locality at
  the level of registers is the target of our optimization
  strategy. Note that the term "(register) locality" is also
  mentioned throughout the paper.  Secondly, we have modified another
  key sentence in the Introduction (page 3):

    "... With such a kernel structure, we focus on aspects like the
     minimization of floating-point operations, register allocation
     and instruction-level parallelism, especially in the form of SIMD
     vectorization."

  into one that we think explicitly targets your concern:

    "... With such kernel and working set structures, traditional
     transformations for cache locality (e.g., blocking) are not
     helpful; instead we focus on aspects like the minimization of
     floating-point operations, register locality and
     instruction-level parallelism, especially vectorization.>>

@@

3) "The current trend for GPGPU is to have the GPU sharing the memory
with the CPU (so called SVM – Shared Virtual Memory). This SVM feature
has already shown up in OpenCL 2.0 and Cuda 6.0 standards. With this
trend, such claims about the stringent memory requirements of GPU will
no longer to be true."

  Even though SVM is going to be part of the new GPU programming
  models (OpenCL, CUDA), that does not mean that "the stringent memory
  requirements will no longer be true." SVM is just an abstraction
  that allows programmers to avoid explictly handling memory
  management; however, behind the scenes, the continuous transfers
  between host and device will still impair the performance. We
  therefore prefer to be really cautious about this matter.

@@

4) "<<There has been some research on adapting local assembly to GPUs
(which we mention in Section 7), although it differs from ours in
several ways, including: (i) not relying on automated code generation
from a domain-specific language (explained next), (ii) testing only
very low order methods, (iii) not optimizing for cross-loop arithmetic
intensity (the goal is rather effective multi-thread
parallelization). >>

Your work focuses on low order methods, yet you claim that similar
work on GPU is different from yours since they work on “very low”
order methods. What are the differences between “low” and “very low”?
From the description of this paper, it seems that they are very
close. Some of the low order methods have trip count s as low as 3, so
how much lower can “very low” go?"

  We definitely agree our description was unclear. Therefore, we
  rewrote that part of the paper as follows:

    "There has been some research on adapting local assembly to GPUs,
     which we mention in Section 7, although it differs from ours in
     several ways, including: (i) not optimizing for cross-loop
     arithmetic intensity (the goal is rather effective multi-thread
     parallelization), and (ii) not relying on automated code
     generation from a domain-specific language (explained next);
     moreover, they have done testing only on order 1 methods, while
     we focus on the order range 1–4."

  This way, we not only remove the previous ambiguity (by explicitly
  stating they only test order 1 methods), but also emphasize the
  importance of points (i) and (ii).

@@

  Fixed

6) "The treatment of expression split is still too simple. Although
real implementations favor simplicity, a technical paper does not."

  For this aspect, we have added a few sentences to Section 3.4, where
  expression splitting is described:

    "... Nevertheless the performance gain from improved register
     reuse can still be greater if suitable heuristics are used. Our
     approach consists of traversing the expression tree and
     recursively splitting it into multiple sub-expressions as long as
     the number of variables independent of the innermost loop exceeds
     a certain threshold. This is elaborated in the next section, and
     validated against empirical search in Section 5.2.6. >>

  This does not make the treatment less simplistic, but helps
  understanding the heuristics we have used (by also adding a forward
  pointer to the upcoming section, in which COFFEE's cost model is
  detailed) and by stating that this heuristics was "validated against
  empirical search." Our performance evaluation shows that this is
  indeed a good heuristic, but as you correctly point out, there is
  space for further systematic exploration. That is why we also
  added a sentence to the Conclusion section, in which we mention
  planned further work:

    "... As suggested by the performance evaluation, an exploration of
     more sophisticated approaches for expression splitting also looks
     promising."

  With the expression splitting transformation, our intention was to
  show that by generalizing the well-known loop fission
  transformation, and through simple intuition, register locality can
  notably be improved. It is clear that a formal study would require
  another complete paper, because the complexity of the problem grows
  considerably if we consider generic expressions embedded in generic
  loop nests. We believe that the ideas provided in this paper
  represent a starting point for an in-depth analysis of the topic.

### 2014-10-22 TACO-paper-v2/coffee.tex
edits

An assembly kernel is characterized by the presence of an affine,
often non-perfect loop nest, in which individual loops are rather
small: their trip count rarely exceeds 30, and may be as low as 3 for
low order methods. In the innermost loop, a problem-specific,
compute intensive expression evaluates a two dimensional array,
representing the result of local assembly in an element of the
discretized domain. With such kernel and working set structures, traditional transformations for cache locality (e.g. blocking) are not helpful; instead we focus on aspects like the minimization of floating-point operations, register locality
and instruction-level parallelism, especially vectorization.

@@

Splitting an expression (henceforth \emph{split}) has, however,
several drawbacks. Firstly, it increases the number of accesses to $A$
in proportion to the ``split factor,'' which is the number of
sub-expressions produced. Also, depending on how splitting is done, it
can lead to redundant computation. For example, the number of times
the product $det*W3[i]$ is performed is proportional to the number of
sub-expressions, as shown in the code snippet. Further, it increases
loop overhead, for example through additional branch
instructions. Finally, it might affect register locality: for
instance, the same array could be accessed in different
sub-expressions, requiring a proportional number of loads be
performed; this is not the case for the Helmholtz
example, though. Nevertheless the performance gain from improved register reuse 
can still be greater if suitable heuristics are used. Our approach 
consists of traversing the expression tree and recursively splitting it into 
multiple sub-expressions if the number of innermost-loop independent 
variables is larger than a certain threshold. This is elaborated in the next section, 
and validated against empirical search in Section~\ref{sec:perf-results-split}.

@@

In this paper, we have presented the study and systematic performance
evaluation of a class of composable cross-loop optimizations for improving
arithmetic intensity in finite element local assembly kernels, and their
integration in a novel compiler, COFFEE. In the context of automated code
generation for finite element local assembly, COFFEE is the first compiler
capable of introducing low-level optimizations to maximize instruction-level
parallelism, register locality and SIMD vectorization. Assembly kernels have
particular characteristics. Their iteration space is usually very small, with
the size depending on aspects like the degree of accuracy one wants to
reach (polynomial order of the method) and the mesh discretization employed.
The data space, in terms of number of arrays and scalars required to evaluate
the element matrix, grows proportionately with the complexity of the
finite element problem. COFFEE has been developed taking into account
all of these degrees of freedom, based on the idea that reducing the
problem of local assembly optimization to a fixed sequence of
transformations is far too superficial if close-to-peak performance
needs to be reached. The various optimizations overcome limitations of
current vendor and research compilers. The exploitation of domain
knowledge allows some of them to be particularly effective, as
demonstrated by our experiments on two state-of-the-art Intel
platforms. Further work includes a comprehensive study about
feasibility and constraints on transforming kernels into a sequence of
calls to external linear algebra libraries. As suggested by the performance evaluation, an analysis of more sophisticated approaches for expression splitting also looks promising. COFFEE supports all of the
problems expressible in Firedrake, and is already integrated with
this framework. The generality and the applicability of the proposed
code transformations to other domains has also been discussed.

>>>>

An assembly kernel is characterized by the presence of an affine,
often non-perfect loop nest, in which individual loops are rather
small: their trip count rarely exceeds 30, and may be as low as 3 for
low order methods. In the innermost loop, a problem-specific, compute
intensive expression evaluates a two dimensional array, representing
the result of local assembly in an element of the discretized
domain. With such kernel and working set structures, traditional
transformations for cache locality (e.g., blocking) are not helpful;
instead we focus on aspects like the minimization of floating-point
operations, register locality and instruction-level parallelism,
especially vectorization.

@@

Splitting an expression (henceforth \emph{split}) has, however,
several drawbacks. Firstly, it increases the number of accesses to $A$
in proportion to the ``split factor,'' which is the number of
sub-expressions produced. Also, depending on how splitting is done, it
can lead to redundant computation. For example, the number of times
the product $det*W3[i]$ is performed is proportional to the number of
sub-expressions, as shown in the code snippet. Further, it increases
loop overhead, for example through additional branch
instructions. Finally, it might affect register locality: for
instance, the same array could be accessed in different
sub-expressions, requiring a proportional number of loads be
performed; this is not the case for the Helmholtz example,
though. Nevertheless the performance gain from improved register reuse
can still be greater if suitable heuristics are used. Our approach
consists of traversing the expression tree and recursively splitting
it into multiple sub-expressions as long as the number of variables
independent of the innermost loop exceeds a certain threshold. This is
elaborated in the next section, and validated against empirical search
in Section~\ref{sec:perf-results-split}.

@@

In this paper, we have presented the study and systematic performance
evaluation of a class of composable cross-loop optimizations for
improving arithmetic intensity in finite element local assembly
kernels, and their integration in a novel compiler, COFFEE. In the
context of automated code generation for finite element local
assembly, COFFEE is the first compiler capable of introducing
low-level optimizations to maximize instruction-level parallelism,
register locality and SIMD vectorization. Assembly kernels have
particular characteristics. Their iteration space is usually very
small, with the size depending on aspects like the degree of accuracy
one wants to reach (polynomial order of the method) and the mesh
discretization employed.  The data space, in terms of number of arrays
and scalars required to evaluate the element matrix, grows
proportionately with the complexity of the finite element
problem. COFFEE has been developed taking into account all of these
degrees of freedom, based on the idea that reducing the problem of
local assembly optimization to a fixed sequence of transformations is
far too superficial if close-to-peak performance needs to be
reached. The various optimizations overcome limitations of current
vendor and research compilers. The exploitation of domain knowledge
allows some of them to be particularly effective, as demonstrated by
our experiments on two state-of-the-art Intel platforms. Further work
includes a comprehensive study about feasibility and constraints on
transforming kernels into a sequence of calls to external linear
algebra libraries. As suggested by the performance evaluation, an
exploration of more sophisticated approaches for expression splitting
also looks promising. COFFEE supports all of the problems expressible
in Firedrake, and is already integrated with this framework. The
generality and the applicability of the proposed code transformations
to other domains has also been discussed.
